\replaced{Fast Fourier transformation of electromagnetic data for computationally expensive kernels}{Fast time-domain electromagnetic modelling in the frequency domain}

### 2020-11-28 ManuscriptTime.tex
Draft of revision to co-authors

% Remove once done
\newcommand{\itodo}[1]{\todo[color=blue!40!white, inline]{\sffamily #1}}
\newcommand{\mtodo}[1]{\todo[inline]{\sffamily #1}}
\newcommand{\etodo}[1]{\todo[color=blue!40!white]{#1}}
\newcommand{\dul}[1]{\underline{\underline{#1}}}
\newcommand{\ul}[1]{\underline{#1}}
\newcommand{\rmk}[1]{{\color{red}{#1}}}
\newcommand{\ohmm}{\ensuremath{\Omega\,}\text{m}\xspace}

@@

\begin{summary}
%
\deleted{Modelling three-dimensional time-domain controlled-source
electromagnetic data with frequen\-cy-domain solvers is an alternative to
time-domain solvers.} \added{Three-dimensional controlled-source
electromagnetic data is often computed directly in the domain of interest,
either in the frequency domain or in the time domain. Computing it in one
domain and transforming it via a Fourier transform to the other domain is a
viable alternative.} It requires the \replaced{evaluation}{computation} of many
\replaced{responses in the computational domain}{frequencies} if standard
Fourier transforms are used, which can make it prohibitively expensive\deleted{
in comparison with a time-domain computation}. The speed of
\replaced{such}{time-domain} modelling \replaced{which is followed by a
transform}{with frequency-domain computation} is defined by three key points:
solver, method and implementation of the Fourier transformation, and gridding.
%
The faster the \deleted{frequency-domain} solver, the faster
\deleted{time-domain} modelling will be. It is important that the solver is
robust over a wide range of \added{values (}frequencies\added{ or times)}.
\deleted{A three-dimensional, iterative, frequency-domain multigrid solver
satisfies this re\-quirement, but any other robust solver will do.}
%
The method should require as few \replaced{kernel evaluations}{frequencies} as
possible while remaining robust. As the frequency \added{and time}
range\added{s} span\deleted{s} many orders of magnitude, the required
\replaced{values}{frequencies} are ideally equally spaced on a logarithmic
scale. The \added{proposed} fast \deleted{frequency-to-time domain} method uses
either the digital linear filter method or the logarithmic fast Fourier
transform together with a careful selection of \replaced{evaluation
points}{frequencies} and interpolation. \replaced{In frequency-to-time domain
tests t}{T}his methodology requires typically 15 to 20 frequencies to cover a
wide range of offsets\deleted{in one-dimensional test models}.
%
The gridding should be frequency-\added{ or time-}dependent, which is
accomplished by making it a function of skin depth. Optimising for the least
number of required cells \deleted{for a given frequency} should be combined
with optimising for computational speed.
%
Looking carefully at these \deleted{three} points results in much smaller
computation times with speedup factors of ten or more over previous results. A
\deleted{frequency-domain} computation \replaced{followed by transformation}{of
time-domain data} can therefore be competitive to \replaced{computation in the
required }{time-}domain \deleted{solvers} if the required \replaced{evaluation
points}{frequencies} and the corresponding grids are carefully chosen.
%
\end{summary}

@@

CSEM codes for 2D and 3D computations, on the other hand, mostly compute their
responses directly in the required domain, either frequency or time. There is a
wealth of 3D electromagnetic codes, with a first boom in the 1990's, with a
good state of the art provided by \cite{B.SEG.99.Oristaglio}. A second wave
occurred during the CSEM boom in the hydrocarbon industry in the early 21th
century, see, e.g., \cite{SG.05.Avdeev} and \cite{SG.10.Borner} for
comprehensive overviews. Recent publications span the width of numerical
methods from integral equation codes \citep{MGS.17.Kruglyakov} and differential
equation methods such as finite differences \citep{CAG.13.Sommer}, finite
elements \citep{GJI.13.Grayver}, and finite volumes \citep{GEO.14.Jahandari}.
Besides the different methods there are also many different types of
discretisation. Commonly used, besides rectilinear Yee grids
\citep{IEEE.66.Yee} as used by our code, are Lebedev grids
\citep{CMMP.64.Lebedev} but also unstructured grids using, for instance,
tetrahedral \citep{CAG.17.Cai}. The computational requirement to model CSEM
data also depends heavily on the technique which is used to solve the system,
where there are direct solvers \citep{GEO.15.Grayver}, indirect solvers
\citep{GJI.15.Jaysaval}, and more recently also hybrid solvers
\citep{GEO.18.Liu}.

Under certain conditions it can be competitive to model time-domain data with a
frequency-domain code, as shown by \cite{GEO.08.Mulder}. The conditions are: a
sufficiently powerful solver, appropriate frequency selection and
interpolation, and an automated gridding, for which they used the
multi-frequency CSEM approach presented by \cite{GEO.07.Plessix}. We build upon
these results but improve the run time from hours to minutes. The main reasons
for this significant speed-up are an adaptive, frequency-dependent gridding
scheme that minimizes the required cells in each dimension, and a logarithmic
Fourier transform such as digital linear filters \citep[DLF, ][]{GP.71.Ghosh}
or the logarithmic fast Fourier transform \citep[FFTLog, ][]{RAS.00.Hamilton}
to go from the frequency to the time domain. The latter makes it also possible
to only use the imaginary part of the frequency-domain response, which has
advantages when it comes to interpolation.

In the next section, we briefly review the methodology as introduced by
\cite{GEO.07.Plessix} and \cite{GEO.08.Mulder} and highlight their advantages
and shortcomings. This is followed by an outline of our changes to the Fourier
transform and the adaptive gridding. Finally, we demonstrate the efficiency of
the approach with some numerical results.

{\color{alizarin}
\verb!v v v REMOVE COMMENTS v v v REMOVE COMMENTS v v v REMOVE COMMENTS v v v!
}
{\color{mygreen}
\begin{itemize}
  \item Too generic; more references
  \item what are the methodological advantages with respect to other
    approaches.
  \item wants: TEM can be accelerated using similar coarse-gridding
    considerations as mentioned in your manuscript. In Commer \& Newman (2005),
    the FDTD method could be accelerated through grid-coarsening during
    runtime, using the fact that the transient field-decay filters out
    high-frequency components.
  \item I do believe the subspace-projection method should definitely be
    discussed in a pro/con manner.
  \item \ldots
\end{itemize}

Dieter: One one hand we agree that the introduction could be expanded, on the
other hand we are not fond of repeating everything again. The introduction
contains quite some extensive, good review papers (Oristaglio and Spies, 1999;
Avdeev, 2005; Börner, 2010; Constable, 2010). The comparison between different
approaches for time-domain computation was done in quite some detail in Mulder
(2008)

TODO: We think that there is a misunderstanding here. First we mention Avdeev
(2005) and Börner (2010), and say that these are very good review papers up to
2005 and 2010, respectively. Recent publications then means publications since
these two review papers were published. We do not want to repeat again all the
pioneering publications, that has been done, but highlight a few recent ones.
We changed the wording in the introduction to make this clear.

Redo intro: Focus on transform, not on frequency- or time-domain EM.

\begin{itemize}
  \item Wim: Include perhaps (one of) explicitly, instead of implicitly:
    Druskin, V. L., and L. A. Knizhnerman, 1994, Spectral approach to solving
    three-dimensional Maxwell’s diffusion equations in the time and frequency
    domains, Radio Science, 29, 937--953. doi: 10.1029/94RS00747 Druskin, V.
    L., L. A. Knizhnerman, and P. Lee, 1999, New spectral Lanczos decomposition
    method for induction modeling in arbitrary 3D geometry, Geophysics, 64,
    701--706. doi: 10.1190/1.1444579.
  \item Evert: I also think we should respond a bit more accommodating the
    reviewer's request. Mulder 2008 is quite old now and papers were published
    after that dealing with solutions directly in time domain. We can write
    that in this paper we include only papers that were not reference in Mulder
    et al. (2008) and refer to that for earlier papers. New are Mittet 2010
    (FDTD of waves and correspondence principle, GPY), Um et al. 2010 (Finite
    element, SEG abstract) and Zhou et al. 2020 TGRS
    (10.1109/TGRS.2020.2979798). In terms of number of references, you can
    decide to not worry to much about that, or remove some others. You have
    several references to works on the differences between measuring/analysing
    data in time or frequency domain. That could go out, e.g.
  \item Evert: i think you can group separate points under the recommendation
    that comes from his first concern. I understand his recommendation as:
    "Write an introduction that is more focused on time domain CSEM modeling."
    That should be done in a comprehensive way and include direct time domain
    methods such as Krylov subspace, Simpeg, FDTD of waves and use
    correspondence principle. It can be interesting to also include more
    arbitrary source time functions and not only step switch on or off and
    impulse response. For the filter approach, in principle, all FD codes are
    suitable to be used. His second concern phrased as a recommendation: "Show
    results from a more complex model." after that the specific points come.
  \item Evert: I think that Commer wanted to refer to Druskin and Knihznerman
    (i am sure spelling is not correct) in radio science 1994, which is most
    likely the older version of the code he mentions. We refer to that 1994
    paper in Mulder 2008.
  \item Evert: The main issue is that the Druskin work it is what Commer uses.
    This is the principal issue. This comes back at point 9. Perhaps now you
    can include it in the introduction as a motivation to work on this topic.
  \item Evert: So let us make an explicit sentence in the introduction to
    clarify this. We also may need to change the text in the introduction to
    avoid making our life difficult by selecting some papers while not
    selecting others. This can be done by only selecting review papers on the
    topic and write that that is our choice here.
  \item Wim: Is that different from doi 10.1190/1.1801936 (mentioned in
    Mulder,Wirianto,Slob)? Then also include Haber from that paper, or just all
    references of section 2 (complexity) in that paper?
  \item Evert: This all belongs to one concern: The introduction is too general
    and insufficiently addresses the actual topic of obtaining time domain data
    by CSEM modeling. That does not mean you need to repeat everything. You
    could start by writing that we will only discuss new developments since
    2008, and refer to Mulder 2008 for older works that relate to the topic of
    time domain CSEM data modeling. I don't think Commer actually wants an
    extensive review of the literature and include his work. He writes that the
    introduction should be more focused on the topic of the paper.
  \item Evert: i have no problem with mentioning in the paper that our idea
    grid coarsening can be used in the method of Commer and Newman (2005). In
    that case no further response is necessary, we tick it off by saying, we
    refer to Commer and Newman (2005) on grid-coarsening. i would limit
    references to Commer to this paper then. I don't think we say anywhere that
    our method is the best in any sense. It seems this reviewer is in
    competition. I don't think we should be in competition. We work for the
    public domain. We put forward our ideas. We leave to the readers to decide
    which options they prefer. We should not be forced need to sell our method.
    Of course, my personal opinion does not change the reality that the AE has
    written that he wants to see the majority of the reviewers comments
    implemented in our revision. We should accept that.
  \item Evert: If we judge this is part of the majorty to be implemented as the
    Associate Editor says, it means the AE agrees with Commer and we should do
    it. No need to add references. In the introduction, just add a summary of
    the order analysis given in Mulder 2008 where some order analysis was
    done.
  \item References provided by Commer:
    \begin{itemize}
      \item Commer \& Newman, 2005, An accelerated time domain finite
        difference simulation scheme for three-dimensional transient
        electromagnetic modeling using geometric multigrid concepts, Radio
        Science, 41, RS3007.
      \item Commer et al., 2006, New results on the resistivity structure of
        Merapi Volcano(Indonesia), derived from three-dimensional restricted
        inversion of long-offset transient electromagnetic data, GJI.
      \item Druskin \& Zaslavsky, 2011, Solution of 3D time-domain
        electromagnetic problems using optimal subspace projection,
        Geophysics.
      \item Weiland, T., 1996, Time domain EM field computation with finite
        difference methods, Int. J. Num. Mod., 9, 295-319.
      \item Yuji Mitsuhata, Toshihiro Uchida, Yutaka Murakami, Hiroshi Amano,
        The Fourier transform of controlled-source time-domain electromagnetic
        data by smooth spectrum inversion, Geophysical Journal International,
        144, January 2001, 123-135.
    \end{itemize}
\end{itemize}

}
{\color{alizarin}
\verb!^ ^ ^ REMOVE COMMENTS ^ ^ ^ REMOVE COMMENTS ^ ^ ^ REMOVE COMMENTS ^ ^ ^!
}

@@

\cite{GEO.08.Mulder} provide a computational complexity analysis of various
methods to model transient electromagnetic responses directly in the
time-domain, and compare it to the computation of transient EM responses in the
frequency-domain with a subsequent Fourier transform. They conclude the review
by stating \emph{\guillemotleft Although it remains to be seen which of the
four methods requires the least computer time for a given accuracy, the
frequency-domain approach appears to be attractive.\guillemotright} Their
\todo{mention the four methods}
approach is to minimize the computation time by having, in addition to the just
introduced adaptive gridding, an adaptive frequency selection scheme. This
scheme starts with computing the responses for a set of just five frequencies,
regularly sampled on a log-scale, from minimum to maximum required frequency.
All the other frequencies are interpolated with a shape-preserving
piecewise-cubic Hermite interpolation \citep[PCHIP, ][]{SIAM.80.Fritsch}.
Testing the stability of the obtained response by removing a single
frequency-value at a time their scheme decides if more frequencies in-between
the already computed ones are required. In this way frequencies are only added
if required, hence if certain criteria of response stability are not met. While
this method is good and effective for a single offset, it loses all its
advantages if one tries to compute different offsets within one computation, as
each offset requires a different set of adaptive frequencies. Additionally, it
hampers the parallelization over frequencies.

@@

For the numerical computations we use the open-source (Apache License 2.0) code
\emg3d \citep{JOSS.19.Werthmuller}, a multigrid solver based on
\cite{GP.06.Mulder}, which can be used as a preconditioner for Krylov subspace
solvers or as a solver on its own. The multigrid approach works fine for the
diffusive approximation of Maxwell's equation, which assumes that
$\omega\varepsilon \ll \sigma$, where $\varepsilon$ is electric permittivity
(F/m). The remaining system to solve in the frequency domain is then given by
the second-order differential equation of the electric field,
%
\begin{equation}
    \mathrm{i}\omega\sigma \mathbf{E} +
    \nabla \times \mu^{-1} \nabla \times \mathbf{E}
    = -\mathrm{i}\omega\mathbf{J}_\mathrm{s} \, ,
  \label{eq:maxwell}
\end{equation}
%
where $\mathbf{E}$ is the electric field (V/m) and $\mathbf{J}_\mathrm{s}$ the
current source (A/m$^2$); time dependence is $\exp(\mathrm{i}\omega t)$. The
standard multigrid approach fails for severe stretching or strong anisotropy,
for which \deleted{\emg3d has implemented }known improvements such as
line-relaxation and semicoarsening \citep{ECCFD.06.Jonsthovel} \added{are
implemented}, with a non-standard Cholesky decomposition to speed up the
computation \citep{GEO.08.Mulder}. One of the big advantages of the multigrid
method is that it scales linearly (optimal) with the grid size in both CPU and
RAM usage \citep{B.Springer.20.Mulder}. This makes it feasible to run even big
models on standard computers, without the need for big clusters. All the
examples in this article are run on a laptop with an i7-6600U\,CPU\,@\,2.6\,GHz
(x4) and 16\,GB of memory, using Ubuntu \replaced{20.04}{18.04} and Python
\replaced{3.8}{3.7}. \added{All examples were run using a single thread, but
parallelization over frequencies is straight forward.} For the semi-analytical
computations of layered models we use the open-source (Apache License 2.0) code
\empymod \citep{GEO.17.Werthmuller}.

@@

For the actual transform we use either the digital linear filter (DLF) method
or the logarithmic fast Fourier transform (FFTLog). The DLF method was
introduced to geophysics by \cite{GP.71.Ghosh}, and is arguably the most common
method in EM geophysics for both its simplicity and its speed. It is
implemented for the Hankel and Fourier transforms in most EM modelling codes,
e.g., \cite{GEO.09.Key}. A simple tool to design digital linear filters was
recently presented by \cite{GEO.19.Werthmuller}, together with a comprehensive
overview of the history and development of DLF in geophysics. FFTLog,
introduced by \cite{RAS.00.Hamilton}, is another transform algorithm which
proved to be powerful for the frequency-to-time domain transformation of EM
responses, e.g., \cite{INT.14.Werthmuller}. In our tests they are both about
equal in speed and accuracy. DLF requires a wider range and many more
frequencies than the FFTLog. Both methods share some important characteristics
in comparison with the standard FFT: the required input frequencies are equally
spaced on a logarithmic scale\added{ (natural logarithm in the case of the DLF
and decimal logarithm in the case of the FFTLog)}, and they only require either
the real or the imaginary part of the frequency-domain response. We can take
advantage of that by using only the imaginary part of the frequency-domain
response. The imaginary part goes to zero when the frequency goes to zero or to
infinity, with the advantage that knowing the endpoints makes it possible to
convert the extrapolation of missing frequencies into \added{an }interpolation.

@@

The example model is a marine scenario with 1\,km water depth of resistivity
$\rho=0.3\,\ohmm$ ($\rho = \sigma^{-1}$) and a 100\,m thick target of
100\,\ohmm at 1\,km below the seafloor in a background of 1\,\ohmm. The source
is 50\,m above the seafloor, the receiver is on the seafloor, and the response
is the inline $x$-directed $E$-field. The left subplot shows the imaginary part
in the frequency domain and the right plot the corresponding impulse response
in the time domain. The red lines are the \replaced{semi-analytical
responses}{precise results obtained from \empymod}. The blue circles indicate
the actually computed responses and the black dots the frequencies which are
interpolated or set to zero. The resulting time-domain response has a relative
error of less than 1\,\% everywhere except for very early times.

Figure \ref{fig:GUI-FFTLog-log} shows exactly the same on a logarithmic scale,
without the interactive GUI-related elements. It can be seen that with the
chosen frequencies the time-domain response starts to divert above about 100\,s
and below 0.3\,s. It also shows the oscillating high-frequency part, which is
hard to interpolate. Figure \ref{fig:GUI-DLF-lin} is the same as
\ref{fig:GUI-FFTLog-lin}, but transformed with the DLF method applying the
81-point sine-cosine filter from \cite{GEO.09.Key}. The same frequencies were
computed as in the FFTLog case and the missing ones interpolated. The error of
the corresponding time-domain response is comparable, so either FFTLog or DLF
can be used.
%
\begin{figure*}
  \centering
  \subfloat[]{{\includegraphics[width=0.5\fwidth]{02a-GUI-FFTLog-log}}%
  \label{fig:GUI-FFTLog-log}}\hfill
  \subfloat[]{{\includegraphics[width=0.5\fwidth]{02b-GUI-DLF-lin}}%
  \label{fig:GUI-DLF-lin}}\\
  \caption{(a) Same as Figure \ref{fig:GUI-FFTLog-lin}, but on a logarithmic
  scale. (b) Using DLF instead of FFTLog, but computing the same
  frequenc\replaced{y range}{ies} as for the FFTLog. The resulting time domain
  response has comparable accuracy (compare the error to the error in Figure
  \ref{fig:GUI-FFTLog-lin}).}
  \label{fig:GUI-log-lin}
\end{figure*}
%
In the above example we used 20 frequencies, but that many would not be
required for the shown offset of 5\,km, a few of the lower and higher
frequencies could be left out. \added{But even so we only need to compute
50\,\% of the required frequencies in the case of the FFTLog, and only 14\,\%
in the case of the DLF.} \replaced{W}{However, w}ith this frequency-selection
we can model a wide range of offsets. This is shown in Figure
\ref{fig:multi-offset}, where the above parameters were used to model the
responses at offsets $r=1.5,3,6$, and 12\,km.
%
\begin{figure*}
  \centering
  \includegraphics[width=0.75\fwidth]{03-multi-offset}
  \caption{Normalized (a) frequency- and (b) time-domain responses for offsets
    $r=1.5,3,6$, and 12\,km using the parameters defined in Figure
    \ref{fig:GUI-FFTLog-lin}. The coloured circles are the actually computed
    responses, the black dots are the responses which are set to zero or
    interpolated. The \replaced{grey}{black} curves are the analytical
    responses.}
  \label{fig:multi-offset}
\end{figure*}
%

@@

For the comparison we use the open-source code \simpeg
\citep{CAG.15.Cockett}\replaced{, which}{. \simpeg} is a framework for modelling
and inversion of geophysical data such as gravity, magnetics, and CSEM. It has
Maxwell's equations implemented in both the frequency and time domain. As such
we can compare our result computed in the frequency domain followed by a
Fourier transform to a result computed directly in the time domain. A principal
difference between \simpeg and \emg3d is that the former has various direct
solvers implemented, whereas the latter is an iterative multigrid solver. The
3D model is therefore a rather small example in order to be able to run it on
our test machine, as the memory requirement by the direct solver would
otherwise be too high. There are not many options out there of open-source
time-domain 3D codes, \simpeg being the one we found to be suitable. A step-off
response was chosen as this is the response currently implemented in
\replaced{the framework}{\simpeg}.

>>>>

% Remove once done
% \newcommand{\itodo}[1]{\todo[color=blue!40!white, inline]{\sffamily #1}}
% \newcommand{\mtodo}[1]{\todo[inline]{\sffamily #1}}
% \newcommand{\etodo}[1]{\todo[color=blue!40!white]{#1}}
\newcommand{\dul}[1]{\underline{\underline{#1}}}
\newcommand{\ul}[1]{\underline{#1}}
\newcommand{\rmk}[1]{{\color{red}{#1}}}
\newcommand{\ohmm}{\ensuremath{\Omega\,}\text{m}\xspace}

@@

\begin{summary}
%
\deleted{Modelling three-dimensional time-domain controlled-source
electromagnetic data with frequen\-cy-domain solvers is an alternative to
time-domain solvers.} \added{Three-dimensional controlled-source
electromagnetic data is often computed directly in the domain of interest,
either in the frequency domain or in the time domain. Computing it in one
domain and transforming it via a Fourier transform to the other domain is a
viable alternative.} It requires the \replaced{evaluation}{computation} of many
\replaced{responses in the computational domain}{frequencies} if standard
Fourier transforms are used\replaced{. This}{, which} can make it prohibitively
expensive \replaced{if the kernel is time-consuming as is the case in
three-dimensional electromagnetic modelling}{in comparison with a time-domain
computation}. The speed of \replaced{such}{time-domain} modelling
\replaced{that is obtained through a transform}{with frequency-domain
computation} is defined by three key points: solver, method and implementation
of the Fourier transformation, and gridding.
%
The faster the \deleted{frequency-domain} solver, the faster
\deleted{time-domain} modelling will be. It is important that the solver is
robust over a wide range of \added{values (}frequencies\added{ or times)}.
\deleted{A three-dimensional, iterative, frequency-domain multigrid solver
satisfies this re\-quirement, but any other robust solver will do.}
%
The method should require as few \replaced{kernel evaluations}{frequencies} as
possible while remaining robust. As the frequency \added{and time}
range\added{s} span\deleted{s} many orders of magnitude, the required
\replaced{values}{frequencies} are ideally equally spaced on a logarithmic
scale. The \added{proposed} fast \deleted{frequency-to-time domain} method uses
either the digital linear filter method or the logarithmic fast Fourier
transform together with a careful selection of \replaced{evaluation
points}{frequencies} and interpolation. \replaced{In frequency-to-time domain
tests t}{T}his methodology requires typically 15 to 20 frequencies to cover a
wide range of offsets\deleted{in one-dimensional test models}.
%
The gridding should be frequency-\added{ or time-}dependent, which is
accomplished by making it a function of skin depth. Optimising for the least
number of required cells \deleted{for a given frequency} should be combined
with optimising for computational speed.
%
Looking carefully at these \deleted{three} points results in much smaller
computation times with speedup factors of ten or more over previous results. A
\deleted{frequency-domain} computation \replaced{followed by transformation}{of
time-domain data} can therefore be competitive to \replaced{computation in the
required }{time-}domain \deleted{solvers} if the required \replaced{evaluation
points}{frequencies} and the corresponding grids are carefully chosen.
%
\end{summary}

@@

CSEM codes for 2D and 3D computations, on the other hand, mostly compute their
responses directly in the required domain, either frequency or time. There is a
wealth of 3D electromagnetic codes, with a first boom in the 1990's, with a
good state of the art provided by \cite{B.SEG.99.Oristaglio}. A second wave
occurred during the CSEM boom in the hydrocarbon industry in the early 21th
century, see, e.g., \cite{SG.05.Avdeev} and \cite{SG.10.Borner} for
comprehensive overviews. \replaced{Focusing on r}{R}ecent publications
\added{after these review articles} span\added{s} the width of numerical
methods from integral equation codes \citep{MGS.17.Kruglyakov} and differential
equation methods such as finite differences
\replaced{\mbox{\citep{GEO.10.Mittet, GEO.11.Zaslavsky, CAG.13.Sommer}}}
{\mbox{\citep{CAG.13.Sommer}}}, finite elements \citep{GJI.13.Grayver}, and
finite volumes \citep{GEO.14.Jahandari}. Besides the different methods there
are also many different types of discretisation. Commonly used, besides
rectilinear Yee grids \citep{IEEE.66.Yee} as used by our code, are Lebedev
grids \citep{CMMP.64.Lebedev} but also unstructured grids using, for instance,
tetrahedral \citep{CAG.17.Cai}. The computational requirement to model CSEM
data also depends heavily on the technique which is used to solve the system,
where there are direct solvers \citep{GEO.15.Grayver}, indirect solvers
\citep{GJI.15.Jaysaval}, and more recently also hybrid solvers
\citep{GEO.18.Liu}.

Under certain conditions it can be \replaced{a viable alternative}{competitive}
to model time-domain data with a frequency-domain code, as shown by
\cite{GEO.08.Mulder}. \added{They compared the computational complexity of
explicit time-stepping schemes (Du Fort-Frankel method as used in, e.g.,
\mbox{\cite{GEO.04.Commer, GEO.07.Maao}}); implicit schemes \mbox{\citep[e.g.,
][]{GEO.04.Haber}}, and matrix exponentials and Lanczos reduction schemes
\mbox{\cite[e.g., ][]{RS.94.Druskin}} with the Fourier transform of
frequency-domain solutions. Their conclusion was that the Fourier transform
method can be favourable if the number of required frequencies is small
relative to the number of grid points in each coordinate.} The conditions
\added{for fast computation of time-domain data with a frequency-domain code}
are: a sufficiently powerful solver, appropriate frequency selection and
interpolation, and an automated gridding, for which they used the
multi-frequency CSEM approach presented by \cite{GEO.07.Plessix}. We build upon
these results but improve the run time from hours to minutes. The main reasons
for this significant speed-up are an adaptive, frequency-dependent gridding
scheme that minimizes the required cells in each dimension, and a logarithmic
Fourier transform such as digital linear filters \citep[DLF, ][]{GP.71.Ghosh}
or the logarithmic fast Fourier transform \citep[FFTLog, ][]{RAS.00.Hamilton}
to go from the frequency to the time domain. The latter makes it also possible
to only use the imaginary part of the frequency-domain response, which has
advantages when it comes to interpolation.

\added{While we use the finite-integration technique
\mbox{\citep{AEU.77.Weiland}} in the frequency domain to show our method it is
important to note that our ideas can be used with any solver, it is not bound
to one or another frequency- or time-domain solver. The ideas for the frequency
selection can be applied to time selection, and time-dependent, adaptive
gridding exists as well \mbox{\citep[e.g., ][]{RS.06.Commer}}.} In the next
section, we briefly review the methodology as introduced by
\cite{GEO.07.Plessix} and \cite{GEO.08.Mulder} and highlight their advantages
and shortcomings. This is followed by an outline of our changes to the Fourier
transform and the adaptive gridding. Finally, we demonstrate the efficiency of
the approach with some numerical results.

@@

\cite{GEO.08.Mulder} provide a computational complexity analysis of various
methods to model transient electromagnetic responses directly in the
time-domain, and compare it to the computation of transient EM responses in the
frequency-domain with a subsequent Fourier transform. They conclude the review
by stating \emph{\guillemotleft Although it remains to be seen which of the
four methods requires the least computer time for a given accuracy, the
frequency-domain approach appears to be attractive.\guillemotright} Their
approach is to minimize the computation time by having, in addition to the just
introduced adaptive gridding, an adaptive frequency selection scheme. This
scheme starts with computing the responses for a set of just five frequencies,
regularly sampled on a log-scale, from minimum to maximum required frequency.
All the other frequencies are interpolated with a shape-preserving
piecewise-cubic Hermite interpolation \citep[PCHIP, ][]{SIAM.80.Fritsch}.
Testing the stability of the obtained response by removing a single
frequency-value at a time their scheme decides if more frequencies in-between
the already computed ones are required. In this way frequencies are only added
if required, hence if certain criteria of response stability are not met. While
this method is good and effective for a single offset, it loses all its
advantages if one tries to compute different offsets within one computation, as
each offset requires a different set of adaptive frequencies. Additionally, it
hampers the parallelization over frequencies.

@@

For the numerical computations we use the open-source (Apache License 2.0) code
\emg3d \citep{JOSS.19.Werthmuller}, a multigrid solver based on
\cite{GP.06.Mulder}, which can be used as a preconditioner for Krylov subspace
solvers or as a solver on its own. The multigrid approach works fine for the
diffusive approximation of Maxwell's equation, which assumes that
$\omega\varepsilon \ll \sigma$, where $\varepsilon$ is electric permittivity
(F/m). The remaining system to solve in the frequency domain is then given by
the second-order differential equation of the electric field,
%
\begin{equation}
    \mathrm{i}\omega\sigma \mathbf{E} +
    \nabla \times \mu^{-1} \nabla \times \mathbf{E}
    = -\mathrm{i}\omega\mathbf{J}_\mathrm{s} \, ,
  \label{eq:maxwell}
\end{equation}
%
where $\mathbf{E}$ is the electric field (V/m) and $\mathbf{J}_\mathrm{s}$ the
current source (A/m$^2$); time dependence is $\exp(\mathrm{i}\omega t)$. The
standard multigrid approach fails for severe stretching or strong anisotropy,
for which \deleted{\emg3d has implemented }known improvements such as
line-relaxation and semicoarsening \citep{ECCFD.06.Jonsthovel} \added{are
implemented}, with a non-standard Cholesky decomposition to speed up the
computation \citep{GEO.08.Mulder}. One of the big advantages of the multigrid
method is that it scales linearly (optimal) with the grid size in both CPU and
RAM usage \citep{B.Springer.20.Mulder}. This makes it feasible to run even big
models on standard computers, without the need for big clusters. All the
examples in this article are run on a laptop with an i7-6600U\,CPU\,@\,2.6\,GHz
(x4) and 16\,GB of memory, using Ubuntu \replaced{20.04}{18.04} and Python
\replaced{3.8}{3.7}. \added{All examples were run using a single thread, but
parallelization over frequencies is straight forward, as the computations are
independent.} For the semi-analytical computations of layered models we use the
open-source (Apache License 2.0) code \empymod \citep{GEO.17.Werthmuller}.

@@

For the actual transform we use either the digital linear filter (DLF) method
or the logarithmic fast Fourier transform (FFTLog). The DLF method was
introduced to geophysics by \cite{GP.71.Ghosh}, and is arguably the most common
method in EM geophysics for both its simplicity and its speed. It is
implemented for the Hankel and Fourier transforms in most EM modelling codes,
e.g., \cite{GEO.09.Key}. A simple tool to design digital linear filters was
recently presented by \cite{GEO.19.Werthmuller}, together with a comprehensive
overview of the history and development of DLF in geophysics. FFTLog,
introduced by \cite{RAS.00.Hamilton}, is another transform algorithm which
proved to be powerful for the frequency-to-time domain transformation of EM
responses, e.g., \cite{INT.14.Werthmuller}. In our tests they are both about
equal in speed and accuracy. DLF requires a wider range and many more
frequencies than the FFTLog. Both methods share some important characteristics
in comparison with the standard FFT: the required input frequencies are equally
spaced on a logarithmic scale\added{ (natural logarithm in the case of the DLF
and decimal logarithm in the case of the FFTLog)}, and they only require either
the real or the imaginary part of the frequency-domain response. We can take
advantage of that by using only the imaginary part of the frequency-domain
response. The imaginary part goes to zero when the frequency goes
\added{either} to zero or to infinity, with the advantage that knowing the
endpoints makes it possible to convert the extrapolation of missing frequencies
into \added{an }interpolation.

@@

The \added{shown} example model is a marine scenario with 1\,km water depth of
resistivity $\rho=0.3\,\ohmm$ ($\rho = \sigma^{-1}$) and a 100\,m thick target
of 100\,\ohmm at 1\,km below the seafloor in a background of 1\,\ohmm. The
source is 50\,m above the seafloor, the receiver is on the seafloor, and the
response is the inline $x$-directed $E$-field. The left subplot shows the
imaginary part in the frequency domain and the right plot the corresponding
impulse response in the time domain. The red lines are the
\replaced{semi-analytical responses}{precise results obtained from \empymod}.
The blue circles indicate the actually computed responses and the black dots
the frequencies which are interpolated or set to zero. The resulting
time-domain response has a relative error of less than 1\,\% everywhere except
for very early times.

Figure \ref{fig:GUI-FFTLog-log} shows exactly the same on a logarithmic scale,
without the interactive GUI-related elements. It can be seen that with the
chosen frequencies the time-domain response starts to divert above about 100\,s
and below 0.3\,s. It also shows the oscillating high-frequency part, which is
hard to interpolate\added{ and we therefore set to zero}. Figure
\ref{fig:GUI-DLF-lin} is the same as \ref{fig:GUI-FFTLog-lin}, but transformed
with the DLF method applying the 81-point sine-cosine filter from
\cite{GEO.09.Key}. The same frequencies were computed as in the FFTLog case and
the missing ones interpolated. The error of the corresponding time-domain
response is comparable, so either FFTLog or DLF can be used.
%
\begin{figure*}
  \centering
  \subfloat[]{{\includegraphics[width=0.5\fwidth]{02a-GUI-FFTLog-log}}%
  \label{fig:GUI-FFTLog-log}}\hfill
  \subfloat[]{{\includegraphics[width=0.5\fwidth]{02b-GUI-DLF-lin}}%
  \label{fig:GUI-DLF-lin}}\\
  \caption{(a) Same as Figure \ref{fig:GUI-FFTLog-lin}, but on a logarithmic
  scale. (b) Using DLF instead of FFTLog, but computing the same
  frequenc\replaced{y range}{ies} as for the FFTLog. The resulting time domain
  response has comparable accuracy (compare the error to the error in Figure
  \ref{fig:GUI-FFTLog-lin}).}
  \label{fig:GUI-log-lin}
\end{figure*}
%
In the above example we used 20 frequencies, but that many would not be
required for the shown offset of 5\,km, a few of the lower and higher
frequencies could be left out. With this frequency-selection we can model a
wide range of offsets. This is shown in Figure \ref{fig:multi-offset}, where
the above parameters were used to model the responses at offsets $r=1.5,3,6$,
and 12\,km. \added{We only need to compute 50\,\% of the required frequencies
in the case of the FFTLog, and only 14\,\% in the case of the DLF.}
%
\begin{figure*}
  \centering
  \includegraphics[width=0.75\fwidth]{03-multi-offset}
  \caption{Normalized (a) frequency- and (b) time-domain responses for offsets
    $r=1.5,3,6$, and 12\,km using the parameters defined in Figure
    \ref{fig:GUI-FFTLog-lin}. The coloured \replaced{symbols}{circles} are the
    actually computed responses, the black dots are the responses which are set
    to zero \added{(high frequencies)} or interpolated \added{(low
    frequencies)}. The \replaced{grey}{black} curves are the analytical
    responses.}
  \label{fig:multi-offset}
\end{figure*}
%

@@

For the comparison we use the open-source code \simpeg
\citep{CAG.15.Cockett}\replaced{, which}{. \simpeg} is a framework for modelling
and inversion of geophysical data such as gravity, magnetics, and CSEM. It has
Maxwell's equations implemented in both the frequency and time domain. As such
we can compare our result computed in the frequency domain followed by a
Fourier transform to a result computed directly in the time domain. A principal
difference between \simpeg and \emg3d is that the former has various direct
solvers implemented, whereas the latter is an iterative multigrid solver. The
3D model is therefore a rather small example in order to be able to run it on
our test machine, as the memory requirement by the direct solver would
otherwise be too high. There are not many options out there of open-source
time-domain 3D codes, \simpeg being the one we found to be suitable. A step-off
response was chosen as this is the response currently implemented in
\replaced{it}{\simpeg}.
